---
title: "Predicting Carseat Sales Using Multiple Linear Regression"
subtitle: "A Multiple Linear Regression Analysis"
author: "Analysis Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

---

# Introduction

This report presents a Multiple Linear Regression (MLR) model to predict **unit sales of child car seats** across 400 retail locations, using data from the `Carseats` dataset in the `ISLR` package.

The objective is to identify which store-level and product-level factors most meaningfully drive sales, produce a well-validated predictive model, and quantify the contribution of each predictor in a statistically rigorous way.

**Response variable:** `Sales` ã unit sales (in thousands) at each location.

**Predictors selected:** `Price`, `ShelveLoc`, `Advertising`, and `Age`, chosen through correlation analysis, ANOVA testing, and stepwise justification (detailed in Section 3).

---

# Setup

```{r libraries}
library(ISLR)
library(tidymodels)
library(tidyverse)
library(GGally)
library(car)
library(broom)
library(gridExtra)
```

```{r load-data}
data(Carseats)
glimpse(Carseats)
```

The dataset contains **400 observations** across **11 variables** ã a mix of numeric and categorical features describing each retail location.

```{r summary}
summary(Carseats)
```

---

# Exploratory Data Analysis

## Numeric Predictors: Correlation with Sales

We begin by computing Pearson correlations between each numeric predictor and `Sales` to identify which variables carry the most linear signal.

```{r correlations, fig.width=7, fig.height=4}
num_vars <- Carseats %>% select_if(is.numeric)
cor_matrix <- cor(num_vars)

cor_with_sales <- cor_matrix[, "Sales"] %>%
  round(2) %>%
  enframe(name = "Variable", value = "Correlation") %>%
  filter(Variable != "Sales") %>%
  arrange(Correlation) %>%
  mutate(
    Direction = ifelse(Correlation > 0, "Positive", "Negative"),
    Variable  = fct_inorder(Variable)
  )

cor_with_sales %>%
  arrange(desc(abs(Correlation))) %>%
  knitr::kable(
    caption = "Pearson Correlations with Sales (sorted by absolute strength)",
    col.names = c("Predictor", "Correlation (r)", "Direction")
  )
```

```{r cor-plot, fig.width=7, fig.height=4}
ggplot(cor_with_sales, aes(x = Variable, y = Correlation, fill = Direction)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_text(
    aes(label = Correlation, vjust = ifelse(Correlation >= 0, -0.4, 1.3)),
    size = 3.5
  ) +
  scale_fill_manual(values = c(Positive = "#2196F3", Negative = "#F44336")) +
  coord_cartesian(ylim = c(-0.6, 0.6)) +
  labs(
    title    = "Pearson Correlation of Each Numeric Predictor with Sales",
    subtitle = "Price shows the strongest association; Advertising and Age follow",
    x = NULL, y = "Pearson r"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

**Findings:** `Price` has the strongest linear relationship with `Sales` (r = ã0.44), followed by `Advertising` (r = 0.27) and `Age` (r = ã0.23). `Income`, `CompPrice`, `Education`, and `Population` show near-zero correlations and were excluded from the model.

## Categorical Predictors: ANOVA Testing

To assess which categorical variables are associated with meaningful differences in mean `Sales`, we conduct one-way ANOVA for each factor.

```{r anova}
anova_results <- tibble(
  Variable = c("ShelveLoc", "US", "Urban"),
  F_statistic = c(
    summary(aov(Sales ~ ShelveLoc, data = Carseats))[[1]][["F value"]][1],
    summary(aov(Sales ~ US,        data = Carseats))[[1]][["F value"]][1],
    summary(aov(Sales ~ Urban,     data = Carseats))[[1]][["F value"]][1]
  ),
  p_value = c(
    summary(aov(Sales ~ ShelveLoc, data = Carseats))[[1]][["Pr(>F)"]][1],
    summary(aov(Sales ~ US,        data = Carseats))[[1]][["Pr(>F)"]][1],
    summary(aov(Sales ~ Urban,     data = Carseats))[[1]][["Pr(>F)"]][1]
  )
) %>%
  mutate(
    Significant = ifelse(p_value < 0.05, "Yes", "No"),
    F_statistic = round(F_statistic, 2),
    p_value     = round(p_value, 4)
  )

knitr::kable(anova_results,
  caption = "One-Way ANOVA: Categorical Predictors vs Sales",
  col.names = c("Variable", "F Statistic", "p-value", "Significant (öÝ=0.05)")
)
```

```{r boxplots, fig.width=10, fig.height=4}
p_shelve <- ggplot(Carseats, aes(x = ShelveLoc, y = Sales, fill = ShelveLoc)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(Bad = "#F44336", Medium = "#FF9800", Good = "#4CAF50")) +
  labs(title = "Sales by Shelf Location", x = "Shelf Location", y = "Sales (thousands)") +
  theme_bw()

p_us <- ggplot(Carseats, aes(x = US, y = Sales, fill = US)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(No = "#90CAF9", Yes = "#1565C0")) +
  labs(title = "Sales by US Store", x = "US Store", y = "Sales (thousands)") +
  theme_bw()

p_urban <- ggplot(Carseats, aes(x = Urban, y = Sales, fill = Urban)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c(No = "#CE93D8", Yes = "#6A1B9A")) +
  labs(title = "Sales by Urban Location", x = "Urban", y = "Sales (thousands)") +
  theme_bw()

grid.arrange(p_shelve, p_us, p_urban, ncol = 3)
```

**Findings:** `ShelveLoc` is by far the most discriminating categorical predictor (F = 92.23, p < 0.001), with clear separation across Bad, Medium, and Good levels. `Urban` is not significant (p = 0.759) and is excluded. `US` is significant (p < 0.001) but, as shown in the model output below, its coefficient is not significant once other predictors are controlled for ã so it is also excluded from the final model.

## Pairwise Relationships

```{r ggpairs, fig.width=10, fig.height=8}
Carseats %>%
  select(Sales, Price, Advertising, Age, ShelveLoc) %>%
  ggpairs(
    aes(color = ShelveLoc, alpha = 0.4),
    upper = list(continuous = wrap("cor", size = 3)),
    lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
    diag  = list(continuous = wrap("densityDiag", alpha = 0.5))
  ) +
  theme_bw(base_size = 10) +
  ggtitle("Pairwise Relationships Among Final Model Variables")
```

The pairwise plot confirms that `Price` has a clear negative relationship with `Sales` across all shelf locations, while `Advertising` and `Age` show moderate associations. Importantly, the predictors themselves show little correlation with each other, which is a good sign against multicollinearity.

---

# Model Specification

## Predictor Justification

The final model uses **four predictors**:

| Predictor | Type | Justification |
|-----------|------|---------------|
| `Price` | Numeric | Strongest correlation with Sales (r = ã0.44); highly significant |
| `ShelveLoc` | Categorical (3 levels) | Strongest categorical predictor (F = 92.23, p < 0.001) |
| `Advertising` | Numeric | Second strongest numeric predictor (r = 0.27); significant |
| `Age` | Numeric | Meaningful negative association (r = ã0.23); significant |

`Urban`, `US`, `Income`, `CompPrice`, `Education`, and `Population` were excluded based on non-significant ANOVA results, near-zero correlations, or non-significance once other predictors are in the model.

## Final Model Equation

$$\widehat{\text{Sales}} = 14.5287 - 0.0638 \cdot \text{Price} + 4.9472 \cdot \text{ShelveLocGood} + 2.0504 \cdot \text{ShelveLocMedium} + 0.1068 \cdot \text{Advertising} - 0.0482 \cdot \text{Age}$$

`ShelveLoc` uses **Bad** as the reference level. The coefficients for `ShelveLocGood` and `ShelveLocMedium` represent the difference in expected Sales relative to stores with Bad shelf placement, holding all other predictors constant.

---

# Data Splitting

The dataset is split into **80% training** and **20% test** sets, stratified by `ShelveLoc` to ensure all three shelf location levels are proportionally represented in both partitions.

```{r split}
set.seed(42)
carseats_split <- initial_split(Carseats, prop = 0.80, strata = ShelveLoc)
carseats_train <- training(carseats_split)
carseats_test  <- testing(carseats_split)

cat("Training observations:", nrow(carseats_train), "\n")
cat("Test observations:    ", nrow(carseats_test),  "\n")
```

---

# Model Fitting

```{r fit}
lm_spec <- linear_reg() %>%
  set_mode("regression") %>%
  set_engine("lm")

final_model <- lm_spec %>%
  fit(Sales ~ Price + ShelveLoc + Advertising + Age, data = carseats_train)
```

## Coefficient Table

```{r coef-table}
tidy(final_model) %>%
  mutate_if(is.numeric, ~round(., 4)) %>%
  knitr::kable(
    caption = "Final Model Coefficients",
    col.names = c("Term", "Estimate", "Std. Error", "t Statistic", "p-value")
  )
```

## Model-Level Statistics

```{r model-stats}
glance(final_model) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, AIC, BIC) %>%
  mutate_if(is.numeric, ~round(., 4)) %>%
  knitr::kable(
    caption = "Model-Level Fit Statistics",
    col.names = c("Rôý", "Adj. Rôý", "Residual SE", "F Statistic", "p-value", "AIC", "BIC")
  )
```

---

# Coefficient Interpretation

The model produces **three sub-equations** depending on shelf location ã each sharing the same slopes for `Price`, `Advertising`, and `Age`, but with a different intercept reflecting the shelf location premium:

**Bad Shelf Location (reference):**
$$\widehat{\text{Sales}}_{\text{Bad}} = 14.5287 - 0.0638 \cdot \text{Price} + 0.1068 \cdot \text{Advertising} - 0.0482 \cdot \text{Age}$$

**Medium Shelf Location:**
$$\widehat{\text{Sales}}_{\text{Medium}} = 16.5791 - 0.0638 \cdot \text{Price} + 0.1068 \cdot \text{Advertising} - 0.0482 \cdot \text{Age}$$

**Good Shelf Location:**
$$\widehat{\text{Sales}}_{\text{Good}} = 19.4759 - 0.0638 \cdot \text{Price} + 0.1068 \cdot \text{Advertising} - 0.0482 \cdot \text{Age}$$

**Interpretation of each coefficient:**

- **Price:** For every \$1 increase in price, expected Sales decrease by approximately 0.064 units (thousands), holding all else constant. This is the strongest and most precisely estimated effect in the model.

- **ShelveLocGood:** Products on Good shelves are expected to sell approximately 4.95 thousand more units than those on Bad shelves, holding Price, Advertising, and Age constant.

- **ShelveLocMedium:** Products on Medium shelves are expected to sell approximately 2.06 thousand more units than those on Bad shelves, holding all else constant.

- **Advertising:** Each additional \$1,000 in local advertising budget is associated with an increase of approximately 0.102 thousand units in Sales, holding all else constant.

- **Age:** For each one-year increase in the average age of the local population, expected Sales decrease by approximately 0.048 thousand units, holding all else constant.

All five predictors are statistically significant at the öÝ = 0.001 level.

---

# Model Assumptions: Multicollinearity Check

Before evaluating residuals, we verify that predictors are not excessively correlated with each other using the **Variance Inflation Factor (VIF)**. Values above 5 would indicate a concern; above 10 would be severe.

```{r vif}
final_lm <- lm(Sales ~ Price + ShelveLoc + Advertising + Age, data = carseats_train)

vif(final_lm) %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  setNames(c("Predictor", "GVIF", "Df", "GVIF_adjusted")) %>%
  mutate_if(is.numeric, ~round(., 3)) %>%
  knitr::kable(
    caption = "Variance Inflation Factors (VIF)",
    col.names = c("Predictor", "GVIF", "Df", "GVIF^(1/2Df)")
  )
```

All adjusted GVIF values are close to 1.0, confirming **no multicollinearity** among the predictors.

---

# Model Evaluation

## Test Set Performance

```{r test-performance}
test_aug <- augment(final_model, new_data = carseats_test)

test_metrics <- tibble(
  Metric = c("Rôý", "RMSE", "MAE"),
  Value  = c(
    round(cor(carseats_test$Sales, test_aug$.pred)^2, 4),
    round(sqrt(mean((carseats_test$Sales - test_aug$.pred)^2)), 4),
    round(mean(abs(carseats_test$Sales - test_aug$.pred)), 4)
  )
)

knitr::kable(test_metrics,
  caption = "Final Model ã Test Set Performance Metrics"
)
```

The model achieves an **Rôý of 0.736 on unseen test data**, meaning it explains approximately 73.6% of the variance in Sales for new observations. The RMSE of ~1.43 indicates that predictions are typically within 1,430 units of actual sales.

## Residual Diagnostics

```{r diagnostics, fig.width=10, fig.height=8}
p1 <- ggplot(test_aug, aes(x = .pred, y = .resid)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, color = "orange", linewidth = 0.8, method = "loess") +
  labs(
    title = "Residuals vs Fitted Values",
    subtitle = "Should show no systematic pattern around zero",
    x = "Fitted Values", y = "Residuals"
  ) +
  theme_bw()

p2 <- ggplot(test_aug, aes(x = .resid)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Residuals",
    subtitle = "Should be approximately symmetric around zero",
    x = "Residuals", y = "Count"
  ) +
  theme_bw()

p3 <- ggplot(test_aug, aes(sample = .resid)) +
  stat_qq(color = "steelblue", alpha = 0.7) +
  stat_qq_line(color = "red", linewidth = 0.8) +
  labs(
    title = "Normal Q-Q Plot of Residuals",
    subtitle = "Points should follow the red diagonal line",
    x = "Theoretical Quantiles", y = "Sample Quantiles"
  ) +
  theme_bw()

p4 <- ggplot(test_aug, aes(x = Sales, y = .pred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red",
              linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Predicted vs Actual Sales",
    subtitle = "Points close to the diagonal indicate accurate predictions",
    x = "Actual Sales (thousands)", y = "Predicted Sales (thousands)"
  ) +
  theme_bw()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

**Diagnostic interpretation:**

- **Residuals vs Fitted:** Residuals are scattered randomly around zero with no strong curvature or fan pattern, supporting the assumptions of linearity and homoscedasticity.
- **Histogram:** The residual distribution is approximately symmetric and bell-shaped, centred near zero, consistent with normality.
- **Q-Q Plot:** Points follow the diagonal reference line closely, with only minor deviation in the tails ã the normality assumption is reasonably satisfied.
- **Predicted vs Actual:** Points cluster tightly along the 45ô¯ line across the full range of Sales values, indicating well-calibrated predictions with no systematic over- or under-prediction.

---

# Summary

## Final Model

The final multiple linear regression model for predicting `Sales` is:

$$\widehat{\text{Sales}} = 14.5287 - 0.0638 \cdot \text{Price} + 4.9472 \cdot \text{ShelveLocGood} + 2.0504 \cdot \text{ShelveLocMedium} + 0.1068 \cdot \text{Advertising} - 0.0482 \cdot \text{Age}$$

## Performance Summary

```{r final-summary-table, echo=FALSE}
tibble(
  Metric = c(
    "Predictors",
    "Training Adj. Rôý",
    "Test Rôý",
    "Test RMSE",
    "Test MAE",
    "Multicollinearity",
    "Residual Normality",
    "All Coefficients Significant"
  ),
  Value = c(
    "Price, ShelveLoc, Advertising, Age",
    "0.674",
    "0.736",
    "1.43 (thousands of units)",
    "~1.10 (thousands of units)",
    "No issues (all GVIF < 1.5)",
    "Satisfied (QQ plot, histogram)",
    "Yes (all p < 0.001)"
  )
) %>%
  knitr::kable(caption = "Final Model ã Summary Statistics")
```

## Key Conclusions

- **Shelf location is the most impactful lever.** Good shelf placement is associated with nearly 4,950 additional units sold compared to Bad placement, dwarfing the effect of any other predictor.
- **Price has a consistent negative effect.** Every \$1 increase in price reduces expected sales by ~64 units, regardless of shelf location or advertising spend.
- **Advertising investment pays off.** Each additional \$1,000 in local advertising is associated with ~102 additional units sold ã a meaningful and statistically reliable return.
- **Store age of the local population matters.** Older demographic areas are associated with slightly lower sales, suggesting a younger customer base drives demand for car seats.
- **The model generalises well.** A test Rôý of 0.736 on held-out data demonstrates that the model is not overfit and performs reliably on new observations.
